---
title: "Bikeshare perdict"
author: "Jiahang Li"
date: "2024-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

```{r library and data, include=FALSE}
library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(gganimate)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(tidycensus)
library(mapview)
library(sfdep)
library(spdep)
library(caret)
library (readr)

options(tigris_class = "sf")
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#7892B5","#8CB9C0","#91B5A9","#EDCA7F","#D98481")
palette4 <- c("#7892B5","#8CB9C0","#EDCA7F","#D98481")
palette2 <- c("#7892B5","#D98481")

tidycensus::census_api_key("e79f3706b6d61249968c6ce88794f6f556e5bf3d", overwrite = TRUE)
```

```{r obtain census geom and bike data, warning=FALSE, message=FALSE, results='hide', cache=TRUE}

bikesharenew <- read.csv(url("https://raw.githubusercontent.com/jiahangl98/MUSA508-A5/main/JC-202403-citibike-tripdata.csv"))


jcCensus <- get_acs(geography = "tract", 
          variables = c("B01003_001", "B19013_001", 
                        "B02001_002", "B08013_001",
                        "B08012_001", "B08301_001", 
                        "B08301_010", "B01002_001"), 
          year = 2021, 
          state = "NJ", 
          geometry = TRUE, 
          county=c("Hudson"),
          output = "wide") %>%
  rename(Total_Pop =  B01003_001E,
         Med_Inc = B19013_001E,
         Med_Age = B01002_001E,
         White_Pop = B02001_002E,
         Travel_Time = B08013_001E,
         Num_Commuters = B08012_001E,
         Means_of_Transport = B08301_001E,
         Total_Public_Trans = B08301_010E) %>%
  select(Total_Pop, Med_Inc, White_Pop, Travel_Time,
         Means_of_Transport, Total_Public_Trans,
         Med_Age,
         GEOID, geometry) %>%
  mutate(Percent_White = White_Pop / Total_Pop,
         Mean_Commute_Time = Travel_Time / Total_Public_Trans,
         Percent_Taking_Public_Trans = Total_Public_Trans / Means_of_Transport)

jcTracts <- jcCensus %>%
  as.data.frame() %>%
  distinct(GEOID, .keep_all = TRUE) %>%
  select(GEOID, geometry) %>% 
  st_sf()
```


```{r add census tract}

bike_census <- st_join(bikesharenew %>% 
                         filter(is.na(start_lat) == FALSE &
                                is.na(start_lng) == FALSE &
                                is.na(end_lat) == FALSE &
                                is.na(end_lng) == FALSE) %>%
                         st_as_sf(., coords = c("start_lng", "start_lat"), crs = 4326),
                       jcTracts %>%  st_transform(crs=4326),
                       join=st_intersects, left = TRUE) %>%
  rename(Origin.Tract = GEOID) %>%
  mutate(start_lng = unlist(map(geometry, 1)),
         start_lat = unlist(map(geometry, 2)))%>%
  filter(!(ride_id %in% c("BEC9CC4D1007C753", "86087F431785EDE7", "A1AAB92631439883", "71938DA085242DCC", "4203C77668C47C75", "C92410CA2AEF3947", "6E02CBF36C90F244", "0259885253FD238E", "B39136B37AFB5FE9", "1EC9376D1559C51C", "CDDF16D3A37BE370"))) %>% 
  as.data.frame()
  
```

```{r distribution of stations}

ggplot()+
  geom_sf(data = jcTracts %>%
          st_transform(crs=4326), fill="white")+
  geom_point(data = bike_census, aes(x=start_lng, y = start_lat), 
            color = "#7892B5", alpha = 1, size = 0.3) + 
  ylim(min(bike_census$start_lat), max(bike_census$start_lat))+
  xlim(min(bike_census$start_lng), max(bike_census$start_lng)) +
  labs(title = "Location of Rides Starting Points in Jersey City") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth=0.8)
        )

```

```{r temporal feature engineering}

bike_census <- bike_census %>%
  mutate(interval60 = floor_date(ymd_hms(started_at), unit = "hour"),
         interval15 = floor_date(ymd_hms(started_at), unit = "15 mins"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>% 
  mutate(time_of_day = case_when(hour(interval60) < 7 | hour(interval60) > 18 ~ "Overnight",
                                 hour(interval60) >= 7 & hour(interval60) < 10 ~ "AM Rush",
                                 hour(interval60) >= 10 & hour(interval60) < 15 ~ "Mid-Day",
                                 hour(interval60) >= 15 & hour(interval60) <= 18 ~ "PM Rush")) %>% 
  mutate(hour = hour(started_at),
                weekend = ifelse(dotw %in% c("Sun", "Sat"), "Weekend", "Weekday"))
```

```{r origin-desintation matrix, warning=FALSE}

lines <- st_sfc(
  lapply(1:nrow(bike_census), function(i) {
    st_linestring(matrix(c(bike_census[i, "start_lng"], bike_census[i, "start_lat"],
                            bike_census[i, "end_lng"], bike_census[i, "end_lat"]), ncol = 2, byrow = TRUE))
  }))

lines_sf <- st_sf(geometry = lines)
lines_sf <- st_set_crs(lines_sf, 4326)

Route <- lines_sf %>% head(200)
mapview(Route, zcol = NULL, color = "#7892B5", linewidth = 0.001, alpha = 0.3) # only show 200
```



# Exploratory Analysis

In this section, we delve into some further exploratory analysis of our data set. 

## Temporal Analysis

Starting with temporal analysis, we first visualized the bikeshare trips per hour in Jersey City in August 2023. It is clear that there's some form of circularity in that there are both peak hours and non-peak hours during a single day. Also, we may see that five consecutive higher peaks are usually followed by two less higher peaks. These correspond to high bikeshare need during rush hours and low bikeshare needs overnight, in early mornings, and over the weekend

```{r bikeshare viz1}
ggplot(bike_census %>%
         group_by(interval60) %>%
         tally())+
  geom_line(aes(x = interval60, y = n), color= "#7892B5")+
  labs(title="Bike Share Trips Per Hour in Jersey City, March 2024",
       x="Date", 
       y="Number of trips") + 
theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth =0.8))
  
```

Then, we visualized the distribution of trips by stations at each hour in August. We may see that the distribution of rides is highly right skewed, with most stations having zero or a few number of trips each hour and a couple of stations having much more rides, up to 40 or even 50 rides at a particular hour on a particular day in August 2023. 

```{r bikeshare viz3, warning=FALSE}

ggplot(bike_census %>%
         group_by(interval60, start_station_name) %>%
         tally())+
  geom_histogram(aes(n), binwidth = 5, fill="#7892B5")+
  labs(title="Bike Share Trips Per Hour by Station in Jersey City, Aug 2023",
       x="Trip Counts", 
       y="Number of Stations") + 
    theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth =0.8))

```

We continue to visualize the distribution of mean number of trips at each station during different times of the day. We recognize that during PM rush, the average number of trips has been the highest throughout, with a couple of stations' average trip exceed 10, implying a relatively high demand for a single station. This is followed by ride demand in mid day, when some stations can have more than 5 rides. It is also surprising to notice that the demand is pretty high overnight, with a few stations having more than 5 rides. 

```{r bikeshare viz2, warning=FALSE, message=FALSE}

bike_census %>%
        group_by(interval60, start_station_name, time_of_day) %>%
         tally()%>%
  group_by(start_station_name, time_of_day)%>%
  summarize(mean_trips = mean(n))%>%
  ggplot()+
  geom_histogram(aes(mean_trips, fill=time_of_day), binwidth = 1)+
  scale_fill_manual(values = palette4, name="Time of Day") + 
  labs(title="Mean Number of Hourly Trips Per Station",
       x="Number of trips", 
       y="Frequency")+
  facet_wrap(~time_of_day) +
theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))

```

After that, we visualized the total number of trips in August separated by day of the week. The two main patterns here are that firstly, trip numbers vary by time of the day. Lowest trip counts occur in early morning while highest trip count occur in late afternoon. Secondly, Tuesday, Wednesday, and Thursday tend to observe higher number of trips. 

```{r bikeshare viz4}

ggplot(bike_census %>% mutate(hour = hour(started_at)))+
     geom_freqpoly(aes(hour, color = dotw), binwidth = 3, lwd = 0.8)+
  scale_color_manual(values = c("#003566","#7892B5","#8CB9C0","#91B5A9","#EDCA7F","#E9B9AA","#D98481"), name = "Day") + 
  labs(title="Bike Share Trips in Jersey City by Day of the Week, March 2024",
       x="Hour", 
       y="Trip Counts") + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))

```

Same type of visualize but this time separated between weekend and weekdays. It is very obvious that we see more trips during weekdays and less trips during weekends. Also, during weekends, number of trips start to increase much later than during weekdays, suggesting the missing of morning rush hour will significantly impact people's use of bikeshare services. 

```{r bikeshare viz 5}

ggplot(bike_census)+
     geom_freqpoly(aes(hour, color = weekend), binwidth = 3, lwd = 1.2)+
  scale_color_manual(values=palette2) + 
  labs(color = "Time of Week") +
  labs(title="Bike Share Trips in Jersey City - Weekend vs Weekday, March 2024",
       x="Hour", 
       y="Trip Counts") + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))

```

Building onto that, we group this dataset by station and time of day to visualize the total number of trips at each station during a specific period of time on weekday or weekend. To simplify our visualization, we selected the top 50 occurrences in our dataset. Note that here, each bar represent an instance of station-time of day-weekday/weekend combination. We are able to see that PM rush hour indeed sees that greatest bikeshare demand for a couple of stations in particular. Weekend rides only appear three times in the top 50 list. Zooming into the detail reveals to us that **Grove St PATH** station observes a total of 1117 rides during PM rush on a weekday, 801 rides overnight on a weekday in August 2023. **Hoboken Terminal - River St & Hudson Pl** station observes a total of 864 rides during PM rush on weekday. **South Waterfront Walkway - Sinatra Dr & 1 St** station observes 319 rides overnight on a weekend and 731 rides during PM rush on a weekday in August 2023. 

```{r bikeshare viz 7, fig.width=10}

to_plot <- bike_census%>% # only show 100 stations
  group_by(start_station_id, start_lat, start_lng, weekend, time_of_day, start_station_name) %>%
  tally() %>% 
  arrange(-n) %>% 
  head(50) 
to_plot$ID <- seq_along(to_plot$start_station_id)

to_plot %>% 
  arrange(-n) %>%
  head(50) %>% 
  ggplot(aes(x = reorder(ID, -n), n, fill = time_of_day, color = weekend)) +
  scale_fill_manual(values = palette4, name="Time of Day") + 
  guides(color="none") + 
  geom_bar(stat = "identity", position="stack") +
  scale_color_manual(values = c("transparent", "black")) +
  labs(title="Top 50 Occurences of Rides By Time")+
  ylab("Number of Rides") +
  xlab("") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))
```

## Spatial Analysis

 

```{r bikeshare viz 6}

ggplot()+
  geom_sf(data = jcTracts %>%
          st_transform(crs=4326), fill = "white")+
  geom_point(data = to_plot,
             aes(x=start_lng, y = start_lat, size=n), color = alpha("#7892B5", 0.7)) + 
  ylim(min(bike_census$start_lat), max(bike_census$start_lat))+
  xlim(min(bike_census$start_lng), max(bike_census$start_lng)) +
  scale_size(range = c(0.5, 8)) + 
  labs(size = "Number of Rides") +
  facet_grid(weekend ~ time_of_day) + 
    labs(title="Locations of Stations with Greatest Bikeshare Demand by Time")+ 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth=0.8)
        )
```


```{r spatial auto cor, warning=FALSE}

moran_data <- bike_census %>%
  group_by(start_station_id, start_lat, start_lng, start_station_name) %>%
  tally() %>% 
  arrange(-n) %>% 
  head(80) %>% 
  st_as_sf(., coords = c("start_lng", "start_lat"), crs = 4326)


coords <-  st_coordinates(moran_data) 
neighborList <- knn2nb(knearneigh(coords, 5))
spatialWeights <- nb2listw(neighborList, style="W")

moranTest <- moran.mc(moran_data$n, 
                      spatialWeights, nsim = 999)

ggplot(as.data.frame(moranTest$res[c(1:999)]), aes(moranTest$res[c(1:999)])) +
  geom_histogram(binwidth = 0.01, fill = "#D98481") +
  geom_vline(aes(xintercept = moranTest$statistic), colour = "#7892B5" ,lwd=1) +
  scale_x_continuous(limits = c(-0.5, 0.5)) +
  labs(title = "Observed and Permuted Moran's I for Stations with Most Rides",
       subtitle = "Observed Moran's I in Red",
       x = "Moran's I",
       y = "Count") +
  theme_light() +   
theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10))
```

```{r spatial cor viz, fig.height=8, fig.width=11}

  ggplot() +
  geom_sf(data = jcTracts, fill = "white") +
  geom_sf(data = moran_data,
          aes(size = n, color = n)) + 
  ylim(min(bike_census$start_lat), max(bike_census$start_lat))+
  xlim(min(bike_census$start_lng), max(bike_census$start_lng)) + 
  scale_size(
    range = c(1, 6),
    guide = guide_legend(
      direction = "horizontal",
      nrow = 1,
      label.position = "bottom")) +
  scale_color_gradientn(colours = hcl.colors(5, "RdBu",rev = TRUE, alpha = 0.9), 
                        guide = guide_legend(direction = "horizontal", nrow = 1)) +
  labs(title = "Stations with Most Rides",
       subtitle =  "Dot size proportional to rides") + 
  theme(legend.position = "bottom") + 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth=0.8)
        )

```



```{r animate map temporal viz, message=FALSE, warning=FALSE, cache=TRUE, fig.show='animate'}
library(gifski)
week13 <-
  filter(bike_census , week == 13 & dotw == "Mon")

week13.panel <-
  expand.grid(
    interval15 = unique(week13$interval15),
    Origin.Tract = unique(bike_census$Origin.Tract))

bike.animation.data <-
  mutate(week13, Trip_Counter = 1) %>%
    right_join(week13.panel) %>% 
    group_by(interval15, Origin.Tract) %>%
    summarize(Trip_Count = sum(Trip_Counter, na.rm=T)) %>% 
    ungroup() %>% 
    left_join(jcTracts, by=c("Origin.Tract" = "GEOID")) %>%
    st_sf() %>%
    mutate(Trips = case_when(Trip_Count == 0 ~ "0 trips",
                             Trip_Count > 0 & Trip_Count <= 3 ~ "1-3 trips",
                             Trip_Count > 3 & Trip_Count <= 6 ~ "4-6 trips",
                             Trip_Count > 6 & Trip_Count <= 10 ~ "7-10 trips",
                             Trip_Count > 10 ~ "11+ trips")) %>%
    mutate(Trips  = fct_relevel(Trips, "0 trips","1-3 trips","4-6 trips",
                                       "7-10 trips","10+ trips"))

bikeshare_animation <-
  ggplot() +
    geom_sf(data = bike.animation.data, aes(fill = Trips)) +
    scale_fill_manual(values = palette5) +
    labs(title = "Bikeshare For One Day in March 2024",
         subtitle = "15 minute intervals: {current_frame}") +
    transition_manual(interval15)  + 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks =element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, linewidth=0.8)
        )
  


anim_save("animation.gif", animation = bikeshare_animation, end_pause = 4, height = 1000, width = 700, fps = 6, renderer = gifski_renderer())




```




## Weather Analysis

On top of space and time, weather can be another significant factor of in predicting bikeshare demand. Generally, we believe bikeshare demand will decrease if it's on a rainy, windy, or a very hot day in our case.  With that in mind, we downloaded the weather data using the `riem_measures` function for Newark airport between Aug 1, 2023 and Sept 1, 2023. Note that the Newark Airport station sufficiently provides temporal weather for all of Jersey City. The weather.Panel is generated to summarize temperature, precipitation, and wind speed for every hour. 

```{r download and wrangle weather data, warning=FALSE}

weather.Panel <- 
  riem_measures(station = "EWR", date_start = "2024-03-01", date_end = "2024-04-01") %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))

```

For the temperature, precipitation, and wind data we collected during this month. We may see that there are a couple of rainy days in August. Wind speed is pretty consistent throughout with a couple of really windy days. Temperature fluctuates on a daily basis with higher temperature during the day and cooler temperature at night. 

```{r visualize weather data, fig.height=11, fig.width=8, warning=FALSE}

grid.arrange(
  ggplot(weather.Panel, aes(interval60,Precipitation)) + geom_line(color="#7892B5") + 
    labs(title="Percipitation", x="Hour", y="Perecipitation") + 
    theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank()), 
  ggplot(weather.Panel, aes(interval60,Wind_Speed)) + geom_line(color="#EDCA7F") + 
    labs(title="Wind Speed", x="Hour", y="Wind Speed") +
    theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank()),  
  ggplot(weather.Panel, aes(interval60,Temperature)) + geom_line(color="#D98481") + 
    labs(title="Temperature", x="Hour", y="Temperature") +theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x=element_text(size=8),
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank()), 
  top="Weather Data ")

```


# Space-Time Panel

In the following steps, we created a study panel where each instance in the panel is a unique combination of space and time. In other words, each row will represent the ride at a particular station during a particular hour. 

```{r create study panel, warning=FALSE, message=FALSE}

study.panel <- 
  expand.grid(interval60=unique(bike_census$interval60), 
              start_station_id = unique(bike_census$start_station_id)) %>%
  left_join(., bike_census %>%
              select(start_station_id, start_station_name, Origin.Tract, start_lng, start_lat)%>%
              distinct() %>%
              group_by(start_station_id) %>%
              slice(1))
```



```{r add info to study panel, warning=FALSE, message=FALSE}

bike.panel <- 
  bike_census %>%
  mutate(Trip_Counter = 1) %>%
  right_join(study.panel) %>% 
  group_by(interval60, start_station_id, start_station_name, Origin.Tract, start_lng, start_lat) %>%
  summarize(Trip_Count = sum(Trip_Counter, na.rm=T)) %>%
  left_join(weather.Panel) %>%
  ungroup() %>%
  filter(is.na(start_station_id) == FALSE) %>%
  mutate(week = week(interval60),
         dotw = wday(interval60, label = TRUE)) %>%
  filter(is.na(Origin.Tract) == FALSE) %>% 
  left_join(., jcCensus %>%
              as.data.frame() %>%
              select(-geometry), by = c("Origin.Tract" = "GEOID"))

```




## Weather Correlation

Now, we could perform more exploratory analysis on our panel data to prepare for our regression model. First, we would like to see how temperature will correlate with the number of bike rides in August. The figure below shows a significant, positive, strong, and linear relationship between temperature and mean trip count for all weeks in August. This means that as the temperature increase, bikeshare demand will increase as well. 

```{r visualize temp, warning=FALSE, message=FALSE}

bike.panel %>%
  group_by(interval60) %>% 
  summarize(Trip_Count = mean(Trip_Count),
            Temperature = first(Temperature)) %>%
  mutate(week = week(interval60)) %>%
  ggplot(aes(Temperature, Trip_Count)) + 
    geom_point(color = "#7892B5") + geom_smooth(method = "lm", se= FALSE, color="#D98481") +
    facet_wrap(~week, ncol=8) + 
    labs(title="Trip Count As a Fuction of Temperature by Week",
         x="Temperature", y="Mean Trip Count") + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))
  

```

Next, we would like to see how wind speed will correlate with the number of bike rides in August. We see that a linear and positive relationship between these two variables, but the strength of such relationship varies between different weeks. This generally means that we see more bikeshare demand during windy days. 

```{r visualize wind, warning=FALSE, message=FALSE}
bike.panel %>%
  group_by(interval60) %>% 
  summarize(Trip_Count = mean(Trip_Count),
            Wind = first(Wind_Speed)) %>%
  mutate(week = week(interval60)) %>%
  ggplot(aes(Wind, Trip_Count)) + 
    geom_point(color = "#7892B5") + geom_smooth(method = "lm", se= FALSE, color="#D98481") +
    facet_wrap(~week, ncol=8) + 
    labs(title="Trip Count As a Fuction of Wind by Week",
         x="Wind Speed", y="Mean Trip Count") + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8)) + 
    theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))
```


## Serial Correlation

Should the number of trips exhibit serial (temporal) correlation, then the time lag features will lead to better predictions. The logic behind is that the number of trips at a particular hour will very much correlate with the number of trips immediately before or after that hour. If we know the number of trips at a particular hour, then it would be easier for us to estimate the number of rides around that time. To achieve that, we calculated 6 different forms of lag hours. 

```{r calculate serial lag}

bike.panel <- 
  bike.panel %>% 
  arrange(start_station_id, interval60) %>% 
  mutate(lagHour = dplyr::lag(Trip_Count,1),
         lag2Hours = dplyr::lag(Trip_Count,2),
         lag3Hours = dplyr::lag(Trip_Count,3),
         lag4Hours = dplyr::lag(Trip_Count,4),
         lag12Hours = dplyr::lag(Trip_Count,12),
         lag1day = dplyr::lag(Trip_Count,24)) %>%
   mutate(day = yday(interval60),hour = hour(interval60))
```

The relationship between these lag hours and number of trips are visualized. We see a strong, linear, and negative relationship between lag12hours and number of trips 12 hours before. This means for example, the number of rides we see at 5pm on a particular day is negatively correlated with the number of rides at 5am that day. In contrast, we see a strong, linear, and positive relationship between laghour and lag1day. This means that the number of rides at a particular hour is highly correlated with the number of rides in the immediate next hour and at the same time the following day. Other than that, the strength of the correlation decreases for lag2hour and lag3hour, meaning that the number of rides at 4pm on a particular day, is more correlated with the number of rides at 5pm, rather than at 6 or 7 pm. 

```{r visualize serial correlations, message=FALSE, warning=FALSE}

as.data.frame(bike.panel) %>%
    group_by(interval60) %>% 
    summarise_at(vars(starts_with("lag"), "Trip_Count"), mean, na.rm = TRUE) %>%
    gather(Variable, Value, -interval60, -Trip_Count) %>%
  ggplot(aes(x = Value , y = Trip_Count)) +
    geom_point(size = 1, color = "#7892B5") + geom_smooth(method = "lm", se= FALSE, color="#D98481") +
  facet_wrap(~Variable, ncol = 6)+
  labs(x = "Variable", y = "Correlation", title = "Correlation Between Serial Lag Trips and Trip Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))

```

## Spatial Correlation

We've seen before that spatial autocorrelation exists here in that the number of trips at a particular station during a particular time is correlated with the number of trips at nearby stations. This means that if we know the number of trips at a station, we may be able to estimate the number of trips at the station next to it. Do achieve that, we created a new dataframe called `lag_result` and wrote a nested loop to loop through each hour in August 2023. For each hour, we then calculate the nearest 5 neighbor of each station and the average of rides for that 5 stations. The results were joined back to our panel dataset. 

```{r calculate spatial lag, message=FALSE, warning=FALSE}
lag_result <- data.frame()

for (d in 61:91) {
  for (h in 0:23) {
    spacelag <- bike.panel %>% 
      filter(day == d & hour == h) %>% 
      st_as_sf(., coords = c("start_lng", "start_lat"), crs = 4326) %>%
      coords.test_lag <-  st_coordinates(spacelag) 
      if(!is.numeric(coords.test_lag)) {
        print("Coordinates are not numeric.")
        print(head(coords.test_lag))  # Print some of the coordinate data to inspect it
        }
      neighborList.test_lag <- knn2nb(knearneigh(coords.test_lag, 5))
      spatialWeights.test_lag <- nb2listw(neighborList.test_lag, style="W")
      spacelag <- spacelag %>% mutate(lagTrip = lag.listw(spatialWeights.test_lag, Trip_Count))
      lag_result <- rbind(lag_result, spacelag)
  }
 
}
```

```{r}
lag_result <- data.frame()

for (d in 61:91) {
  for (h in 0:23) {
    spacelag <- bike.panel %>%
      filter(day == d & hour == h) %>%
      st_as_sf(., coords = c("start_lng", "start_lat"), crs = 4326)

    if (!exists("spacelag")) {
      print(paste("spacelag not created for day", d, "hour", h))
      next 
    }

    coords.test_lag <-  st_coordinates(spacelag) 
    coords.test_lag <- is.numeric(coords.test_lag)
    if(!all(sapply(coords.test_lag, is.numeric))) {
      print("Not all coordinates are numeric:")
      print(coords.test_lag[!sapply(coords.test_lag, is.numeric)])
      }
    neighborList.test_lag <- knn2nb(knearneigh(coords.test_lag, 5))
    spatialWeights.test_lag <- nb2listw(neighborList.test_lag, style="W")
    spacelag <- spacelag %>% mutate(lagTrip = lag.listw(spatialWeights.test_lag, Trip_Count))

    lag_result <- rbind(lag_result, spacelag)
  }
}
```

```{r}
print(class(spacelag))
```

```{r calculate spatial lag, message=FALSE, warning=FALSE}
bike.panel <- lag_result %>% 
  mutate(start_lng = unlist(map(geometry, 1)),
         start_lat = unlist(map(geometry, 2))) %>% 
 st_drop_geometry() 
  
  
```

The figure below shows that for all weeks in August, there's a strong, linear, and positive relationship between the number of trips and the number of trips in nearby stations. We will therefore include this variable in our following regression analysis. 

```{r visualize spatial correlation, message=FALSE, warning=FALSE}

bike.panel %>%
  group_by(interval60) %>% 
  summarize(Trip_Count = mean(Trip_Count),
            Lag_Count = mean(lagTrip)) %>%
  mutate(week = week(interval60)) %>%
  ggplot(aes(Lag_Count, Trip_Count)) + 
    geom_point(size = 1, color = "#7892B5") + geom_smooth(method = "lm", se= FALSE, color="#D98481") +
    facet_wrap(~week, ncol=8) + 
    labs(title="Correlation Between Spatial Lag Trips and Trip Count",
         x="Lag Trip Count", y="Mean Trip Count") + 
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"), 
        axis.text.x = element_blank(),
    axis.ticks.x = element_blank(), 
        axis.text.y=element_text(size=8), 
        axis.title=element_text(size=10), 
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.8))

```